---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# inlamesh3d

<!-- badges: start -->
<!-- badges: end -->

The goal of inlamesh3d is to add 3D mesh functionality to INLA.
When the methods have been tested, they will be incorporated into the main
package instead.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("finnlindgren/inlamesh3d", build_vignettes = TRUE)
```
## Example

You must load `INLA` before loading `inlamesh3d` to get the correct overrides.

```{r example}
library(INLA)
library(inlamesh3d)
loc <- matrix(rnorm(12), 4, 3)
tv <- geometry::delaunayn(loc)
mesh <- inla.mesh3d(loc, tv)
str(mesh)
```
New versions of `inla.spde2.matern`, `inla.spde.make.A` and `inla.mesh.fem`
are now available to use in much the same way as for 2D meshes.

For a full example, see `vignette("spde3d")`.
