---
title: "SPDE models on 3D meshes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SPDE models on 3D meshes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `inlamesh3d` package is a temporary add-on package for `INLA` to test new
code for using tetrahedralisation meshes for 3D SPDE models. Load `INLA` first,
so that the new methods provided by `inlamesh3d` take precedence:

```{r setup}
library(INLA)
library(inlamesh3d)
```

# A simple test case

To test the basic features, we construct a mesh with a single tetrahedron:
```{r}
mesh <- inla.mesh3d(rbind(c(0, 0, 0),
                          c(1, 0, 0),
                          c(0, 1, 0),
                          c(0, 0, 1)),
                    rbind(c(1, 2, 3, 4)))
```
The finite element structure matrices can be obtained in the same way as for
the traditional 2D triangle meshes:
```{r}
fem <- inlamesh3d::inla.mesh.fem(mesh)
```

We can extract the matrices to construct a precision matrix manually, for
$(1-\nabla\cdot\nabla)u(s)=\dot{W}(s)$:
```{r}
Q <- with(fem, c0 + 2 * g1 + g2)
solve(Q)
```
On a larger 3D domain, this model has an exponential covariance function, but
with just a single tetrahedron, and Neumann boundary conditions, we should not
expect to see this in this test case.

The new `param2.matern` method is used to construct the covariance parameter
prior distribution model, which is then used by the new version of
`inla.spde2.matern`:
```{r}
spde <- inlamesh3d::inla.spde2.matern(
  mesh,
  param = param2.matern(mesh,
                        alpha = 2,
                        prior_range = 0.1,
                        prior_sigma = 1))
```
The internal parameterisation used by `param2.matern` is different to the old
`param2.matern.orig` parameterisation, and is more similar to the one used by
`inla.spde2.pcmatern`, in that the user-visible aspects are range and standard
deviation instead of $\tau$ and $\kappa$.

Mapping data locations works the same as before. Here we verify that mapping the mesh vertices themselves produces an identity matrix:
```{r}
A <- inlamesh3d::inla.spde.make.A(mesh, mesh$loc)
A
```

# Estimation example

```{r}
mesh_n <- 1000
mesh_loc <- matrix(rnorm(mesh_n * 3), mesh_n, 3)
tetra <- geometry::delaunayn(mesh_loc)
mesh <- inla.mesh3d(loc = mesh_loc, tv = tetra)
range0 <- 1 # Prior median range
sigma0 <- 1 # Prior median standard deviation
spde <- inlamesh3d::inla.spde2.matern(
  mesh,
  param = param2.matern(mesh,
                        alpha = 2,
                        prior_range = range0,
                        prior_sigma = sigma0))
```

The parameterisation is in log-range and log-sigma relative to the prior medians, so that
`theta = log(c(range, sigma) / c(range0, sigma0))` is used to specify the internal parameter vector:
```{r}
Q <- inla.spde2.precision(spde, theta = log(c(1.5, 2) / c(range0, sigma0)))
A0 <- inlamesh3d::inla.spde.make.A(mesh, cbind(0,0,0))
D0 <- rowSums(mesh$loc^2)^0.5

ggplot2::ggplot(data.frame(dist = D0,
                           covariance = as.vector(solve(Q, t(A0))),
                           exponential = 2^2 * exp(-D0 * (sqrt(8 * 0.5) / 1.5)))) +
  ggplot2::geom_point(ggplot2::aes(dist, covariance)) +
  ggplot2::geom_line(ggplot2::aes(dist, exponential), col = "red")
```
We can see some boundary effect due to the finite domain but overall the covariances follow the theory.

Let's simulate a field and observe at some locations (here chosen as a subset of the mesh points):
```{r}
x <- inla.qsample(1, Q)[, 1]
loc <- mesh$loc[seq_len(100), , drop = FALSE]
Aobs <- inlamesh3d::inla.spde.make.A(mesh, loc)
data <- data.frame(y = as.vector(Aobs %*% x) + rnorm(nrow(loc), sd = 0.1))
```

```{r}
stk <- inla.stack(data = list(y = data$y),
                  A = list(Aobs),
                  effects = list(inla.spde.make.index("field", spde$n.spde)))
est <- inla(y ~ -1 + f(field, model = spde),
            data = inla.stack.data(stk, spde = spde),
            control.predictor = list(A = inla.stack.A(stk)),
            verbose = FALSE)
```
Extracting the parameter estimates and transforming them to user-interpretable
scale:
```{r}
quant <- c("0.025quant", "0.5quant", "0.975quant")
param_estimates <- rbind(
  range = exp(est$summary.hyperpar["Theta1 for field", quant]) * range0,
  sigma = exp(est$summary.hyperpar["Theta2 for field", quant]) * sigma0,
  obs_sigma = as.vector(as.matrix(
    est$summary.hyperpar["Precision for the Gaussian observations", rev(quant)]^-0.5
  ))
)
```
```{r, echo=FALSE}
knitr::kable(param_estimates)
```
The observation noise standard deviation is underestimated (true value was $0.1$), but the range and sigma parameters recover the true values, $1.5$ and $2$, respectively.

# Plotting

Experimental code:

```{r eval=FALSE}
plot(mesh)
```

```{r eval=FALSE}
locgrid <- as.matrix(expand.grid(1:10, 1:11, 1:12))+rnorm(10*11*12,sd=1e-6)
meshgrid <- inla.mesh3d(locgrid, geometry::delaunayn(locgrid))
plot(meshgrid,
     include = c(0.5, 0.5, 1),
     alpha=c(0.5, 0.5, 0.9),
     fill_col = inla.generate.colors(meshgrid$loc)$colors)
rgl::clipplanes3d(rbind(c(1,0,0), c(0,1,0), c(0,0,1)), d = 0)
```

# PC prior

Using a PC prior implementation.
For illustration purposes, we use priors with the true parameter values as prior
medians for the spatial range (`range`) and field standard deviation (`sigma`).

```{r}
spde2 <- inlamesh3d::inla.spde2.pcmatern(
  mesh,
  alpha = 2,
  prior.range = c(range0, 0.5),
  prior.sigma = c(sigma0, 0.5)
)
est2 <- inla(y ~ -1 + f(field, model = spde2),
            data = inla.stack.data(stk, spde2 = spde2),
            control.predictor = list(A = inla.stack.A(stk)),
            verbose = FALSE)
param_estimates2 <- rbind(
  range = est2$summary.hyperpar["Range for field", quant],
  sigma = est2$summary.hyperpar["Stdev for field", quant],
  obs_sigma = as.vector(as.matrix(
    est2$summary.hyperpar["Precision for the Gaussian observations", rev(quant)]^-0.5
  ))
)
```
```{r, echo=FALSE}
knitr::kable(param_estimates2)
```
